Quantizing the Bosonic String on a Loop Quantum Gravity Background

### 2024-06-22 lqg-strings.tex
made some changes in the introduction and wrote the paragraphs about inver metric Polyakov, quantum KG string and emergent dimensions

In the decades prior to the advent of string theory a great deal of effort was put into attempting to quantize gravity via both the Lagrangian and Hamiltonian approaches. The Hamiltonian approach involves, as stated earlier, a choice of the foliation of the background geometry into a collection of spacelike surfaces $\Sigma_t$. Central to this approach was the ADM (Arnowitt-Deser-Misner) formalism which allows one to construct a Hamiltonian $H_{GR}$ for general relativity starting from the Einstein-Hilbert action $S_{EH}$. The resulting phase space is co-ordinatized by a configuration variable which is the 3-metric $h_{ab}$ of the ``leaves'' $\Sigma_t$ of the foliation and a momentum variable which is a function of a the extrinsic curvature $k_{ab}$ of the leaves\footnote{Our notational convention is the following. Lowercase Greek letters from the middle of the alphabet $\mu, \nu, \ldots$ are spacetime indices which run from $(0 \ldots 3)$, whereas lowercase Roman letters $a,b,\ldots$ are spatial indices for quantities which live solely on the surfaces $\Sigma_t$ and take values in $(1,2,3)$}. The gravity Hamiltonian turns out to be a sum of two constraints known as the diffeomorphism (or ``momentum'') constraint $H_{diff}$ and the Hamiltonian constraint $H_{ham}$:
\begin{equation}
	H_{GR} = H_{diff} + H_{ham},
\end{equation}
and these are, in turn, functionals on the phase space of general relativity written in terms of $h_{ab}$ and $k_{ab}$. The idea then is that physical states of the theory $\ket{\Psi_{phys}} $, which can be written as functionals $\Psi[h_{ab}]$ of the 3-metric $h_{ab}$ , must be annihilated by these constraints:
\begin{equation}
	H_{GR} \ket{\Psi_{phys}} \equiv 0
\end{equation}
While this procedure is straightforward in principle, in practice is was impossible to implement in the quantum theory due to highly complicated non-polynomial dependence of the diffeomorphism and Hamiltonian constraints on the configuration and momentum variables. This unfortunate state of affairs persisted until the late 1980s when Abhay Ashtekar \cite{Ashketar1986New,Ashtekar1987New} recast general relativity as a theory of a Minkowski tetrad $E^I_\mu$ and a self-dual $\sltwoc$ connection $A_\mu^{IJ}$. Ashtekar realised that the constraints of general relativity, when expressed in terms of these ``new variables'', simplified drastically and became polynomial functions of the tetrad and connection variables. It was quickly realised by various researchers that the resulting theory could be quantized using the same methods used to quantize Yang-Mills gauge theory \cite{Ashtekar1992Weaving,Ashtekar2004Background} and that the diffeomorphism constraint could be solved exactly in terms of so-called \emph{spin-network} states \cite{Rovelli1988Knot,Rovelli1990Loop,Ashtekar1992Gravitons}.

Following this, work by Rovelli and Smolin \cite{Rovelli1993Area,Rovelli1994Discreteness} demonstrated a remarkable feature of this theory, which came to be known as ``Loop Quantum Gravity'', was that one could construct quantum operators for geometric quantities such as areas of two dimensional surfaces and volumes of three dimensional regions. Moreover, these operators could be diagonlized exactly in the spin-network basis and a lower bound on the smallest possible quantum of area and quantum of volume could be derived. This was the first time that physicists had discovered the ``atoms of space'' or the ``quanta of geometry'' in the true sense of the expression, in a theory of quantum gravity. One of the early triumphs of this formalism, which has unfortunately not received enough attention over the years by the larger high energy theory community, was the calculation of the entropy of a Schwarzschild black hole \cite{Rovelli1996Black} in a fully quantum mechanical framework and showing that it matched the expected expression $S \sim A/4 $ for the Bekenstein-Hawking entropy.

One of the primary criticisms of string theory is its apparent lack of ``background independence''. This is more general than the principle of \emph{general covariance} which states that physics should be independent of the choice of co-ordinates. Background independence is the statement that any \emph{quantum} theory of gravity should provide a description of physics which is independent, not only of the choice of co-ordinates, but also of the choice of background manifold on which those co-ordinates live. Such a theory should provide a description not only of classical spacetimes, but also of fluctuating, semiclassical geometries and those in the deep quantum regime which do not have any sensible description in terms of any Riemannian geometry. On the face of it, string theory, at least in its conventional form, does not satisfy this principle. The great virtue of LQG on the other hand, is that it respects background independence as a sacred principle and proceeds to quantize gravity while ensuring that (a notion of) background independence forms the core of the quantum theory.

@@

Before proceeding let us state our ``central dogma''. We take as physically well-motivated the central result of LQG that at the Planck scale well-defined geometric operators - such as those which measure the area and volume of a given region of space - exist, have discrete spectra and are gapped, \ie, the smallest eigenvalue is greater than zero. We now wish to understand how to construct a worldsheet action which will reflect this central dogma.

The following is the outline of the paper. In \autoref{sec:modified-ng} we recall the Nambu-Goto action and suggest a simple modification which incorporates our central dogma, which is that \emph{there must exist a smallest possible area $\Delta$ which is physically measurable.} \footnote{In recent work \cite{Borissova2024From} a very similar expression to ours for the modified Nambu-Goto action has been arrived at by considering a version of LQG called ``area metric gravity''.} We show that this action can be expanded in terms of a small coupling constant $g = l_{pl}/l_s$ where the Planck length - the length at which quantum geometric effects become important - $l_{pl}$ is much smaller than the string scale $l_s$ and therefore $g \ll 1$. We point out that the lowest order correction to the NG action can be viewed as a form of bimetric gravity. In \autoref{sec:bimetric-string} we expand upon this observation and suggest a new action for the ``bimetric string'', along the lines of bimetric gravity. We find both, the Nambu-Goto and the Polaykov forms of this action. Further we show that the phenomenological constant $\Delta$ introduce earlier can be viewed as the modulus of Kalb-Ramond field pulled back to the worldsheet.

In \autoref{sec:inverse-metric-Polyakov} we construct the Polyakov version of the modified NG action, show that the equations of motion it yields are the same as those coming from the modified NG action. Further we solve the string equations of motion to find the fundamental modes and perform the quantization of the bosonic string in the usual manner. In \autoref{sec:quantum-kg-string} we ...

In \autoref{sec:holographic-strings} we describe how to construct an action for the bosonic string which encodes the geometry of the bulk spacetime in the language of the connection formulation of general relativity. We discuss possible avenues for quantizing this action and further studying its implications for the relationship between string theory and LQG.

There are various appendices containing details of calculations and background information. In Appendix \ref{sec:modified-eom} we derive the equations of motion for the modified Nambu-Goto action. Appendix \ref{sec:relation-with-NG-case} shows how the Nambu-Goto and Polakov versions of the modified string action are equivalent to each other.

>>>>

In the decades prior to the advent of string theory a great deal of effort was put into attempting to quantize gravity via both the Lagrangian and Hamiltonian approaches. The Hamiltonian approach involves, as stated earlier, a choice of the foliation of the background geometry into a collection of spacelike surfaces $\Sigma_t$. Central to this approach was the ADM (Arnowitt-Deser-Misner) formalism which allows one to construct a Hamiltonian $H_{GR}$ for general relativity starting from the Einstein-Hilbert action $S_{EH}$. The resulting phase space is coordinatized by a configuration variable which is the 3-metric $h_{ab}$ of the ``leaves'' $\Sigma_t$ of the foliation and a momentum variable which is a function of the extrinsic curvature $k_{ab}$ of the leaves\footnote{Our notational convention is the following. Lowercase Greek letters from the middle of the alphabet $\mu, \nu, \ldots$ are spacetime indices which run from $(0 \ldots 3)$, whereas lowercase Roman letters $a,b,\ldots$ are spatial indices for quantities which live solely on the surfaces $\Sigma_t$ and take values in $(1,2,3)$}. The gravity Hamiltonian turns out to be a sum of two constraints known as the diffeomorphism (or ``momentum'') constraint $H_{diff}$ and the Hamiltonian constraint $H_{ham}$:
\begin{equation}
	H_{GR} = H_{diff} + H_{ham},
\end{equation}
and these are, in turn, functionals on the phase space of general relativity written in terms of $h_{ab}$ and $k_{ab}$. The idea then is that physical states of the theory $\ket{\Psi_{phys}} $, which can be written as functionals $\Psi[h_{ab}]$ of the 3-metric $h_{ab}$ , must be annihilated by these constraints:
\begin{equation}
	H_{GR} \ket{\Psi_{phys}} \equiv 0
\end{equation}
While this procedure is straightforward in principle, in practice it was impossible to implement in the quantum theory due to the highly complicated non-polynomial dependence of the diffeomorphism and Hamiltonian constraints on the configuration and momentum variables. This unfortunate state of affairs persisted until the late 1980s when Abhay Ashtekar \cite{Ashketar1986New,Ashtekar1987New} recast general relativity as a theory of a densitized Minkowski tetrad
\begin{equation} \label{eqn:densitized-tetrad}
	\widetilde{E}^{\mu} _I = \frac{1}{\sqrt{- g}} e^{\mu} _I
\end{equation}
and the self-dual part of the $\sltwoc$ spin-connection
\begin{equation} \label{eqn:self-dual-spin-connection}
	A_{\mu} ^{I J} = \omega ^{I J} _{\mu} - \frac{i}{2} \varepsilon ^{I J} _{\ \ K L} \omega ^{K L} _{\mu} .
\end{equation}
Ashtekar realized that the constraints of general relativity, when expressed in terms of these ``new variables'', simplified drastically and became polynomial functions of the tetrad and connection variables. It was quickly realized by various researchers that the resulting theory could be quantized using the same methods used to quantize Yang-Mills gauge theory \cite{Ashtekar1992Weaving,Ashtekar2004Background} and that the diffeomorphism constraint could be solved exactly in terms of so-called \emph{spin-network} states \cite{Rovelli1988Knot,Rovelli1990Loop,Ashtekar1992Gravitons}.

Following this, work by Rovelli and Smolin \cite{Rovelli1993Area,Rovelli1994Discreteness} demonstrated a remarkable feature of this theory, which came to be known as ``Loop Quantum Gravity'', was that one could construct quantum operators for geometric quantities such as areas of two dimensional surfaces and volumes of three dimensional regions. Moreover, these operators could be diagonlized exactly in the spin-network basis and a discrete spectrum for quantum of area and quantum of volume could be derived. This was the first time that physicists had discovered the ``atoms of space'' or the ``quanta of geometry'' in the true sense of the expression, in a theory of quantum gravity. One of the early triumphs of this formalism, which has unfortunately not received enough attention over the years by the larger high energy theory community, was the calculation of the entropy of a Schwarzschild black hole \cite{Rovelli1996Black} in a fully quantum mechanical framework and showing that it matched the expected expression $S \sim A/4$ for the Bekenstein-Hawking entropy.

One of the primary criticisms of string theory is its apparent lack of ``background independence''. This is more general than the principle of \emph{general covariance} which states that physics should be independent of the choice of co-ordinates. Background independence is the statement that any \emph{quantum} theory of gravity should provide a description of physics which is independent, not only of the choice of co-ordinates, but also of the choice of background manifold on which those co-ordinates live. Such a theory should provide a description not only of classical spacetimes, but also of fluctuating, semiclassical geometries and those in the deep quantum regime which do not have any sensible description in terms of any Riemannian geometry. On the face of it, string theory, at least in its conventional form, does not satisfy this principle. The great virtue of LQG on the other hand, is that it respects background independence by construction and proceeds to quantize gravity while ensuring that (a notion of) background independence forms the core of the quantum theory.

@@

Before proceeding let us state our ``central dogma''. We take as physically well-motivated the central result of LQG that at the Planck scale well-defined geometric operators - such as those that measure the area and volume of a given region of space - exist and have discrete spectra, thus the transition from no area/volume to some area/volume is a discrete jump rather than a continuous one. We now wish to understand how to construct a worldsheet action which reflects this central dogma.

The following is the outline of the paper. In \autoref{sec:modified-ng} we recall the Nambu-Goto action and suggest a simple modification which incorporates our central dogma, which is that \emph{there must exist a smallest possible area $\Delta$ which is physically measurable.} \footnote{In recent work \cite{Borissova2024From} a very similar expression to ours for the modified Nambu-Goto action has been arrived at by considering a version of LQG called ``area metric gravity''.} We show that this action can be expanded in terms of a small coupling constant $k = l_{pl} / l_s$ where the Planck length - the length at which quantum geometric effects become important - $l_{pl}$ is much smaller than the string scale $l_s$ and therefore $k \ll 1$. We point out that the lowest order correction to the NG action can be interpreted in the context of bimetric gravity. In \autoref{sec:bimetric-string} we expand upon this observation and suggest a new action for the ``bimetric string'', along the lines of bimetric gravity. We find both, the Nambu-Goto and the Polaykov forms of this action. Furthermore, we show that the phenomenological correction $\Delta$ introduced earlier is related to the pullback of the Kalb-Ramond field $B_{\mu \nu}$ into the worldsheet.

In \autoref{sec:inverse-metric-Polyakov} we explore the Polyakov version of the modified NG action derived from the bimetric analysis and point out its consequences for the equations of motion. Furthermore, we solve the string equations of motion for a particular case of constant-coefficient to find the fundamental modes and derive the Poisson bracket algebra for the modes and constraints. In \autoref{sec:quantum-kg-string} we follow the route of canonical covariant quantization for the solution previously found and discuss the consequences of our area-corrected model regarding the dimensionality of the background spacetime and the presence of negative-norm states (ghosts) in the physical excitations.

In \autoref{sec:emergent-dimensions} we complement Afshordi's work \cite{Afshordi-emergent} with insights from LQG, showing that LQG quantum geometry naturally constrains most of the free parameters of said model, leading to a more concrete theory. We also suggest a possible consequence for the parameters that remain unconstrained and a generalization of the model.

In \autoref{sec:holographic-strings} we describe how to construct an action for the bosonic string which encodes the geometry of the bulk spacetime in the language of the connection formulation of general relativity. We discuss possible avenues for quantizing this action and further study its implications for the relationship between string theory and LQG.

There are various appendices containing details of calculations and background information. In Appendix \ref{sec:modified-eom} we derive the equations of motion for the modified Nambu-Goto action. Appendix \ref{sec:relation-with-NG-case} shows how the equations of motion extracted from the Nambu-Goto and Polakov versions of the modified string action are equivalent to each other.
